---
title: "Class 12: Genome Informatics"
author: 'Zoe Matsunaga (PID: A16853288)'
date: "2025-05-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion of G/G in a population

Downloaded a CSV File from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880;sample=MXL#373531_tablePanel <

Here we read the CSV file we obtained:

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population, GBR, "great Britian."

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find portion of GIG

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) *100,2)
```

This varient associated with childhood asthma is more prevelant in the GBR population than the MXL population.

Let's dig deeper into this!

# Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

> Q13. How many samples do we have? What is their median expression levels for each of these geneotypes?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

Sample size per genotype:

```{r}
table(expr$geno)
```

Median Expression Levels for each genotype:

```{r}
library(dplyr)

summary_expr <- expr %>%
  group_by(geno) %>%
  summarise(
    median_expression = median(exp, na.rm=TRUE)
  )

print(summary_expr)
```

> Q14. Generate a boxplot with a box per genotype.

```{r}
library(ggplot2)

ggplot(expr) + aes(geno, exp, fill=geno) + 
  geom_boxplot(notch=TRUE) +
  geom_jitter(alpha=0.5, width=0.2) + 
  labs(title="Expression per Genotype",x="Genotype", y="Expression")
 
```

> Q14 continued. What could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

A|A has a higher median expression than G|G displayed in the plot. Yes, the SNP does effect the expression of ORMDL3, with the A allese suggesting higher association with ORMDL3. 

